Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-29 manuscript.rmd
Changed hydraulic traits organ label from upper to lower case

- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data. Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{Leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.

@@

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~Stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

@@

\
Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{x,~Leaf}$ vs. $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen as it best-fitted all but one species based on AIC criteria (Fig. S\@ref(fig:obs-plc-curves)), described as

\begin{equation}
K_{Leaf} = Ae^{-B\Psi_{Leaf}} + error
(\#eq:vc-exp)
\end{equation}

@@

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs. $\\Psi_{x,~Leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

\
\begin{equation}
K_{Leaf,z,n} = Ae^{-B\Psi_{z,n}}
(\#eq:k-soil)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} + error
(\#eq:M1)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,n} StdVPD_{n} + error
(\#eq:M2)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy} + error
(\#eq:M3)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~Leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

@@

```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~Stem}$) in two angiosperm systems. High Kmax is associated with photosynthetic rates with faster growth potential.

@@

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized *K*~Leaf~ vs. $\\Psi$~Leaf~, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on *K*~Leaf~ vs. $\\Psi$~Leaf~ for ", sp.soft.filled$AB.data.model," species vs. those that were predicted from models based on species soft-traits, WSG and LMA.")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_data_vs_model.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/B_data_vs_model.jpeg", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```

>>>>

- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data. Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.

@@

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

@@

\
Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen as it best-fitted all but one species based on AIC criteria (Fig. S\@ref(fig:obs-plc-curves)), described as

\begin{equation}
K_{leaf} = Ae^{-B\Psi_{leaf}} + error
(\#eq:vc-exp)
\end{equation}

@@

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

\
\begin{equation}
K_{leaf,z,n} = Ae^{-B\Psi_{z,n}}
(\#eq:k-soil)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} + error
(\#eq:M1)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,n} StdVPD_{n} + error
(\#eq:M2)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy} + error
(\#eq:M3)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

@@

```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems. High Kmax is associated with photosynthetic rates with faster growth potential.

@@

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for ", sp.soft.filled$AB.data.model," species vs. those that were predicted from models based on species soft-traits, WSG and LMA.")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_data_vs_model.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/B_data_vs_model.jpeg", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```
